---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
We must first set our working directory, unzip the data, and then load it.
```{r echo=TRUE}
setwd("~/GitHub/RepData_PeerAssessment1")
unzip("activity.zip")
ActivityData<-read.csv(file="activity.csv")
```

For much of what is to come, I intend to use `dplyr` to manipulate the data, and so I load it now and convert the dataframe to a `tbl_df` which will allow us to get a nice preview of the data as well.

```{r message=FALSE}
library(dplyr)
ActivityData<-tbl_df(ActivityData)
ActivityData
```
We see there are three variables, `steps,date,interval` with lots of `NA`.

## What is mean total number of steps taken per day?
We start by calculating the mean total number of steps per day. To for ease of understanding, we will make use of `dplyr` to `summarise` the data grouped by `date` to first find the total in each day, which we will store in a new `data.frame` called `DailyTotals`.

```{r echo=TRUE}
DailyTotals<-ActivityData %>%
     group_by(date) %>%
     summarise(total=sum(steps,na.rm=TRUE))
DailyTotals
```

Now that we have the `DailyTotals` we can simply take the mean.

```{r echo=TRUE}
DailyMean<-mean(DailyTotals$total)
DailyMean
```
We see that the mean number of total steps taken each day is `r DailyMean`

In order to get a feeling for how the number of steps vary daily, we will make a histogram of the `DailyTotals$total` data. In addition, we add a vertical line at the mean value.
```{r echo=TRUE}
hist(DailyTotals$total)
abline(v=DailyMean)
```

The mean of a dataset can be greatly influenced by days of abnormally high or abnormally low data values. As a result, we will also consider the median of this data.

```{r echo=TRUE}
DailyMedian<-median(DailyTotals$total)
DailyMedian
```

In this case, we see that the median of `r DailyMedian` is a bit higher than the mean of `r DailyMean`.

## What is the average daily activity pattern?
People tend to do different things throughout the day. To investigate how many steps, on average, occurs throughout the day, we compute the daily average per time interval. Again, `dplyr` is useful to `group_by` the `interval` and take a mean. We preview the data as well.
```{r echo=TRUE}
MeanAtIntervals<-ActivityData %>%
     group_by(interval) %>%
    summarise(MeanSteps=mean(steps,na.rm=TRUE))
MeanAtIntervals
```
For a visual display, we will make a time series plot.

```{r echo=TRUE}
plot(x=MeanAtIntervals$interval,y=MeanAtIntervals$MeanSteps,type="l")
```

By looking at the plot, we see there is a failry clearly maximum average point which we will now locate.

```{r echo=TRUE}
MaxMeanInterval<-MeanAtIntervals[which.max((MeanAtIntervals$MeanSteps)),]
MaxMeanInterval
```

We determine the maximum mean steps is `r MaxMeanInterval$MeanSteps` which occurs at the interval `r MaxMeanInterval$interval`.


## Imputing missing values
Our original dataset contains lots of missing values for steps. In fact, few can calculate how many exactly. We will simply count the number of rows in the subset of data that have at least one missing value.

```{r echo=TRUE}
NumNAs<-nrow(ActivityData[!complete.cases(ActivityData)])
NumNAs
```
From this calculatuation, we see a total of `r NumNAs` contain missing values.


If we assume that, on average, a person's distribution of steps is somewhat consistent each day, we can impute the missing values using the values we do know. We first make use of the dataset we already calculated called `MeanAtIntervals`.


There is a really cool function called `ifelse` that we can use to create a new value `ImputedSteps` which is either the actual steps or the average steps based on `NA`.

```{r echo=TRUE}
ImputedActivityData<-ActivityData %>%
     mutate(ImputedSteps=ifelse(is.na(steps),MeanAtIntervals$MeanSteps[interval],steps))

```

Lets examine the output of this by comparing the dataset `ImputedActivityData` with `ActivityData` first by looking at 5 of the cases where we knew the value.

```{r echo=TRUE}
ImputedActivityData[complete.cases(ActivityData),][650:655,]
```

In this case, we see the `steps` value is equal to the `ImputedSteps`, which is a requirement for imputing. Now let us look at some which required imputation.

```{r echo=TRUE}
ImputedActivityData[!complete.cases(ActivityData),][33:38,]
```

Unforunately, we see that this new dataset also has some missing values shown as `Na`. This corresponds to the fact that there are some intervals for which we have no value on any day. So we will need to impute a second time, ensuring we do have a value. For simplicity, we will simply use the mean of the `steps` across our entire dataset we certainly know exists.

```{r echo=TRUE}
MeanSteps<-mean(ActivityData$steps,na.rm=TRUE)
MeanSteps
ImputedActivityData<-ImputedActivityData %>%
     mutate(ImputedSteps=ifelse(is.na(ImputedSteps),MeanSteps,ImputedSteps))
ImputedActivityData[!complete.cases(ActivityData),][33:38,]
```

We displayed the same data where we say missing values before. We see these missing values are now filled in with the `MeanSteps` value of `r MeanSteps`. Let us confirm there are no missing values by counting the number of missing values similar to how we did earlier.

```{r echo=TRUE}
NumImputedNAs<-nrow(ImputedActivityData[is.na(ImputedActivityData$ImputedSteps),])
NumImputedNAs
```

We are happy to see this value is now `r NumImputedNAs`, indicating full imputation is done.

We will recalulate new daily totals, mean, and median for comparison. Notice this time we need not use the command `is.na()`

```{r echo=TRUE}
NewDailyTotals<-ImputedActivityData %>%
     group_by(date) %>%
     summarise(total=sum(ImputedSteps))
NewDailyTotals
```

Now that we have the `NewDailyTotals` we can simply take the mean.

```{r echo=TRUE}
NewDailyMean<-mean(NewDailyTotals$total)
NewDailyMean
```

Similarly, we can take the median.

```{r echo=TRUE}
NewDailyMedian<-median(NewDailyTotals$total)
NewDailyMedian
```

We can easy see now that the former mean of `r DailyMean` has been replaced by the new mean of `r NewDailyMean`, which is higher. Moreover, the former median of `r DailyMedian` has been replaced by the new median of `r NewDailyMedian` which is also higher. Note that the new values are displayed in scientific notation.

Let's graph a histogram of the new values too.
```{r echo=TRUE}
hist(NewDailyTotals$total)
abline(v=NewDailyMean)
```

This new histogram shows the data has now shifted a bit toward the mean which makes sense since we used the means to replace missing values, making the mean more meaningful. (See what I did there? `HaHa`)


## Are there differences in activity patterns between weekdays and weekends?

People often exhibit different activity on weekends and weekdays. Lets do a simple exploration into this. First we will create a factor variable indicating if the date is a weekday or a weekend. We use the complete imputed data set for this.

```{r echo=TRUE}
ImputedActivityData <- ImputedActivityData %>%
     mutate(day=weekdays(as.Date(date,'%Y-%m-%d')),wkday=ifelse(day %in% c("Saturday", "Sunday"), "weekend", "weekday"))
ImputedActivityData
```

I will now aggregate the data based on these designations.
```{r echo=TRUE}
WkndVsWkd<-ImputedActivityData %>%
     group_by(wkday,interval) %>%
     summarise(AvgSteps=mean(ImputedSteps))
WkndVsWkd
```

I wil make use of the `lattice` package to plot the patterns on these two factors.
```{r echo=TRUE}
library(lattice)
xyplot(AvgSteps ~ interval | wkday, data = WkndVsWkd,type="l")
```

Based on these graphs, you see the data does indeed seem to be different. The weekend actitivy seems to be fairly consistent, whereas on the weekday there seems to be more of an early morning spike.

Thank you for reading my project.

